### 2018-04-01 paper/sexSelectionSpacing-ver1.tex
Edit Conclusion

At one extreme, women without education mostly follow the standard 
pattern of shorter spacing when a woman does not have the desired number of 
son and with limited evidence of biased sex ratios and the increases in
spacing that should follow from the use of sex selection.
 In the situations where there are elevated sex ratios for women without
education, the spacing patterns are close to those with more regular sex
ratios. 
Hence, it is difficult to conclude that the changes in sex
ratios indicate sex selection rather than normal variation or problems
with recall errors.

At the other extreme, women with eight or more years of education have
experienced an almost complete reversal of their spacing patterns.
Rather than having the next birth sooner as they mostly did before 
sex selection became available, women with either no or one son now 
have substantially longer spacing than if they have two or more sons.
As shown by the increasingly significant deviations in the sex ratio
from the natural sex ratio, these changes are likely due to increased
use of sex selection rather than changes in son preference. 
Banning prenatal sex determination has therefore not been able to
reverse the use, although it is possible that it has slowed the rate 
of increase.

Over the last 40 years, there has been a general increase in median
spacing for all education groups, but especially for the most 
educated women.
The exception is for the spacing between marriage and first birth
where there are two counteracting forces: women have become less likely 
to conceive before marriage, and there has been a substantial compression 
in the variation of spacing.
The result is that median spacing from marriage to first birth
that has changed little or even declined for the best-educated women.
Both the decrease for the first spell and the increase for higher-order
spells are consistent with an increased access to contraceptives over
time \citep{Yeakey2009}.

The empirical model is designed to allow for the effects of covariates
to change within the spell. Non-proportionality is essential because the
use of sex selection may vary within a spell. The use of sex selection
may change either because preferences for not too long spacing between
births override the preference for sons or because of selection, where
the women who remain in a spell are those with the strongest son
preference who have been through multiple pregnancies because the
fetuses were female. Because these move in opposite directions, I cannot
directly establish the extent to which they affect the results. Both
are, however, present, as can be seen from the cases where there is
either a decline in sex ratio with spell duration or a substantial
increase with spell duration. The first case is more likely at lower
parities where there is still less pressure to ensure a son, which
allows any preference for shorter spacing to play more of a role. The
second is on the other hand more likely to have an effect for higher
parities where the desire to limit further childbearing and ensure a son
dominates any preferences for shorter spacing.

The potential for changes in sex ratio within a spell means
that censoring is an important issue when trying to understand the 
development in sex ratio.
This is especially the case when it comes to establishing whether 
the use of sex selection is increasing or decreasing.
Take, for example, the case of women with eight or more years of
education in the last period. 
If we look at women who started their third spell in either the
2005--2009 or the 2010--2014 periods and we treat end of 2015
as the cut-off time, then the observed sex ratio for the 2005--2009
women is 55.99\% boys while it is 55.38\% for women who started 
in the 2010--2014 period.%
\footnote{
This combines urban and rural women; the basis result 
changes little if I split by area of residence.
I chose the 2014 cut-off to ensure that both groups cover 
five years, even though NFHS-4 took place in 2015 and 2016.
}
It would, therefore, appear that the use of sex selection for
this spell is slightly decreasing over time.
This interpretation, however, ignores the role of censoring.
If we use the estimation results to predict final sex ratio
once childbearing for the third spell is over, the predicted
sex ratios are 56.10\% boys and 56.08\% boys for the 
2005--2009 and the 2010--2015 starts.
The reason for the difference is that while for the 
2005--2009 start we would have observed 7,134 birth out
of a predicted 7,451 total number of births, we would
only have observed 3,348 births out of a predicted 8,842
for the 2010--2014 group.

Although it is not possible to directly observe whether sex selection
causes the unequal sex ratio and longer spacing the results are
consistent with a general increase in sex selection over time. 
The apparent increase in the use of sex selection is mainly found among
women with more education and more so in urban than in rural areas,
although the differences between urban and rural areas are relatively
minor. 
Compared to previous research one result stands out. 
It appears that the use of sex selection is spreading to women with lower 
levels of education so that even women with one to seven years of 
education show unequal sex ratios and longer spacing for their higher-order spells.

The results here point to two important, interconnected, questions 
for future research:
What is the connection between falling fertility and the 
use of sex selection and is the use of sex selection increasing 
or decreasing over time?
It is clear that fertility has been declining over time in India,
although it is important to note that the decline might be
overstated precisely because of increased use of sex selection and
the resulting longer spacing between births makes it less likely 
that we will observe births to women, even though they eventually
will happen.
The fact that biased sex ratio occurs at lower-order spells over
time and is starting to show up for women with less education 
all indicate that lower fertility and use of sex selection are
related, but a more detailed analysis is needed.
Furthermore, although examining spells separately, as done here, is a useful
first step in understanding how sex selection changes over time,
it does not lend itself easily to examine the degree to which
sex selection changes over time.
For that, a combined model that incorporates the full fertility
history of women would be more appropriate.
A model of this type would allow us to jointly model fertility
and sex selection and help address both questions.

The results here also lead to a set of broader questions that
future research should address.
First, to what extent are the improvements in health for girls
relative to boys the result of selection, the longer spacing between
births, or changing son preference?
There is some early evidence that sample selection can make it appear that
girls are healthier, even though the underlying cause is a combination
of sex selection and higher mortality together with recall 
error \citep{Portner2018a}. 
Understanding what role these factors play in better health for
girls is especially important when evaluating policies that 
aim to directly limit the use of sex selection rather than 
changing preferences or incentives.
If the better health outcomes for girls are, for example, an unintended 
side-effect of the longer spacing that arises from sex-selective 
abortions then an effective ban on sex selection may, at least
temporarily, worsen health outcomes for girls.

Second, we need to understand how female labor force participation
interacts with the use of sex selection.
Increased autonomy for women, for example from better opportunities
for working outside the home, has been suggested as a way to increase
women's status and thereby lower the use of sex selection \citep{Das-Gupta2016}.
This may, however, be a double-edged sword.
On the one hand, it is an clear benefit to the women who gain  
bargaining power, and it increases the cost of repeated 
sex-selective abortions because the increased duration between births
would cause a stronger disruption in labor market participation.
On the other hand, it may further lower desired fertility, and 
everything else equal, that may lead to higher use of sex selection.
Understanding the trade-off between long-term benefits from improvements in
women's labor force participation and short-term costs from potential 
increases in sex selection is of paramount importance.

>>>>

At one extreme, women without education mostly follow the standard 
pattern of shorter spacing when a woman does not have the desired number of 
sons.
There is also limited evidence of biased sex ratios and the increases in
spacing that should follow from the use of sex selection.
 In the situations where there are elevated sex ratios for women without
education, the spacing patterns are close to those with more regular sex
ratios. 
Hence, it is difficult to conclude that the changes in sex
ratios indicate sex selection rather than normal variation or problems
with recall errors.

At the other extreme, women with eight or more years of education have
experienced an almost complete reversal of the traditional spacing patterns.
Rather than having the next birth sooner as they mostly did before 
sex selection became available, women with either no or one son now 
have substantially longer spacing than if they have two or more sons.
As shown by the increasingly significant deviations in the sex ratio
from the natural sex ratio, these changes are likely due to increased
use of sex selection rather than changes in son preference. 
Banning prenatal sex determination has not been able to
reverse the use, although it is possible that it has slowed the rate 
of increase.

Over the last 40 years, there has been a general increase in median
spacing for all education groups, but especially for the most 
educated women.
The exception is for the spacing between marriage and first birth
where there are two counteracting forces: women have become less likely 
to conceive before marriage, and there has been a substantial compression 
in the variation of spacing.
The result is that median spacing from marriage to first birth
has changed little or even declined for the best-educated women.
Both the decrease for the first spell and the increase for higher-order
spells are consistent with increased access to contraceptives over
time \citep{Yeakey2009}.

The empirical model allows for the effects of covariates
to vary within the spell. 
Non-proportionality is essential because the use of sex selection may 
vary within a spell. 
The use of sex selection may change either because preferences for not 
too long spacing between births override the preference for sons or 
because of selection, where
the women who remain in a spell are those with the strongest son
preference who have been through multiple pregnancies because the
fetuses were female. 
Because these move in opposite directions, I cannot
directly establish the extent to which they affect the results. 
Both are, however, present, as can be seen from the cases where there is
either a decline in sex ratio with spell duration or a substantial
increase with spell duration. 
The first case is most likely at lower
parities where there is still less pressure to ensure a son, which
allows any preference for shorter spacing to play more of a role. 
The second is most likely for higher
parities where the desire to limit further childbearing and ensure a son
dominates any preferences for shorter spacing.

The potential for changes in sex ratio within a spell means that
censoring is an important issue when trying to understand how the sex
ratio changes over time, and especially when it comes to establishing
whether the use of sex selection is increasing or decreasing.
Take, for example, the case of women with eight or more years of
education in the last period. 
If we look at women who started their third spell in either the
2005--2009 or the 2010--2014 periods and we treat end of 2015
as the cut-off point, then the observed sex ratio for the 2005--2009
women is 55.99\% boys while it is 55.38\% for women who started 
in the 2010--2014 period.%
\footnote{
The calculations combines urban and rural women; the basis result 
changes little if I split by area of residence.
I chose the 2014 cut-off to ensure that both groups cover 
five years, even though NFHS-4 took place in 2015 and 2016.
}
It would, therefore, appear that the use of sex selection for
this spell is slightly decreasing over time.
This interpretation, however, ignores the role of censoring.
If we use the estimation results to predict final sex ratio
once childbearing for the third spell is over, the predicted
sex ratios are 56.10\% boys and 56.08\% boys for the 
2005--2009 and the 2010--2015 starts, which indicates
virtually no change over time.
The reason for the difference is that while we would have 
observed 7,134 birth out of a predicted 7,451 total number 
of births for the 2005--2009 start, we would only have observed 
3,348 births out of a predicted 8,842 for the 2010--2014 group.

Although it is not possible to directly observe whether sex selection
causes the unequal sex ratio and longer spacing the results are
consistent with a general increase in sex selection over time. 
The apparent increase in the use of sex selection is mainly found among
women with more education and more so in urban than in rural areas,
although the differences between urban and rural areas are relatively
minor. 
Compared to previous research one result stands out. 
It appears that the use of sex selection is spreading to women with lower 
levels of education.
Even women with one to seven years of education show unequal sex ratios 
and longer spacing for their higher-order spells.

The results here point to two important, interconnected, questions 
for future research:
What is the connection between falling fertility and the 
use of sex selection and is the use of sex selection increasing 
or decreasing over time?
It is clear that fertility has been declining over time in India,
although it is important to note that the decline might be
overstated precisely because of increased use of sex selection and
the resulting longer spacing between births makes it less likely 
that we will observe births to women, even though they eventually
will happen.
The fact that biased sex ratio occurs at lower-order spells over
time and is starting to show up for women with less education 
all indicate that lower fertility and use of sex selection are
related, but a more detailed analysis is needed.
Furthermore, although examining spells separately, as done here, is a useful
first step in understanding how sex selection changes over time,
it does not lend itself easily to examine the degree to which
sex selection changes over time.
For both, a combined model that incorporates the full fertility
history of women would be more appropriate.

The results here also lead to a set of broader questions that
future research should address.
First, to what extent are the improvements in health for girls
relative to boys the result of selection, the longer spacing between
births, or changing son preference?
There is some early evidence that sample selection can make it appear that
girls are healthier, even though the underlying cause is a combination
of sex selection and higher mortality together with recall 
error \citep{Portner2018a}. 
Understanding what role these factors play in better health for
girls is especially important when evaluating policies that 
aim to directly limit the use of sex selection rather than 
changing preferences or incentives.
If the better health outcomes for girls are, for example, an unintended 
side-effect of the longer spacing that arises from sex-selective 
abortions, then an effective ban on sex selection may, at least
temporarily, worsen health outcomes for girls.

Second, we need to understand how female labor force participation
interacts with the use of sex selection.
Increased autonomy for women, arising, for example, from better opportunities
for working outside the home, has been suggested as a way to increase
women's status and thereby lower the use of sex selection \citep{Das-Gupta2016}.
This may, however, be a double-edged sword.
On the one hand, it is a clear benefit to the women who gain  
bargaining power, and it increases the cost of repeated 
sex-selective abortions because the increased duration between births
would cause a stronger disruption in labor market participation.
On the other hand, it may further lower desired fertility, and 
that may, everything else equal, lead to higher use of sex selection.
Understanding the trade-off between long-term benefits from improvements in
women's labor force participation and short-term costs from potential 
increases in sex selection is of paramount importance.
